---
title: "Introduction to R, using data"
author: "Data Carpentry contributors"
---

```{r, echo = FALSE, purl = FALSE, message = FALSE}
source("setup.R")
```

------------

> ### Learning Objectives
>
> * Define the following terms as they relate to R: object, assign, call,
>   function, arguments, options.
> * Create objects and assign values to them  in R.
> * Learn how to _name_ objects.
> * Save a script file for later use.
> * Use comments to inform script.
> * Solve simple arithmetic operations in R.
> * Call functions and use arguments to change their default options.
> * Load external data from a .csv file into a data frame.
> * Explore data.frame using commands in R.

------------

You can get output from R simply by typing math in the console:

```{r, purl = FALSE}
3 + 5
12 / 7
```

### Functions and their arguments

You can also use functions 
```{r, purl = FALSE}
sqrt(9)
round(3.14159)
```
Functions are "canned scripts" that automate more complicated sets of commands
including operations assignments, etc. Many functions are predefined, or can be
made available by importing R *packages* (more on that later). A function
usually takes one or more inputs called *arguments*. Functions often (but not
always) return a *value*. A typical example would be the function `sqrt()`. The
input (the argument) must be a number, and the return value (in fact, the
output) is the square root of that number. Executing a function ('running it')
is called *calling* the function. An example of a function call is:

```{r, eval = FALSE, purl = FALSE}
sqrt(100)
```

Here, the `sqrt()` function calculates the square root, and returns the value. This function is very simple, because it takes just one argument.

The return 'value' of a function need not be numerical (like that of `sqrt()`),
and it also does not need to be a single item: it can be a set of things, or
even a dataset. We'll see that when we read data files into R.

Arguments can be anything, not only numbers or filenames, but also other
objects. Exactly what each argument means differs per function, and must be
looked up in the documentation (see below). Some functions take arguments which
may either be specified by the user, or, if left out, take on a *default* value:
these are called *options*. Options are typically used to alter the way the
function operates, such as whether it ignores 'bad values', or what symbol to
use in a plot.  However, if you want something specific, you can specify a value
of your choice which will be used instead of the default.

Let's try a function that can take multiple arguments: `round()`.

```{r, results = 'show', purl = FALSE}
round(3.14159)
```

Here, we've called `round()` with just one argument, `3.14159`, and it has
returned the value `3`.  That's because the default is to round to the nearest
whole number. If we want more digits we can see how to do that by getting
information about the `round` function.  We can use `args(round)` to find what 
arguments it takes, or look at the
help for this function using `?round`.

```{r, results = 'show', purl = FALSE}
args(round)
```

```{r, eval = FALSE, purl = FALSE}
?round
```

We see that if we want a different number of digits, we can
type `digits = 2` or however many we want.

```{r, results = 'show', purl = FALSE}
round(3.14159, digits = 2)
```

If you provide the arguments in the exact same order as they are defined you
don't have to name them:

```{r, results = 'show', purl = FALSE}
round(3.14159, 2)
```

And if you do name the arguments, you can switch their order:

```{r, results = 'show', purl = FALSE}
round(digits = 2, x = 3.14159)
```

It's good practice to put the non-optional arguments (like the number you're
rounding) first in your function call, and to then specify the names of all optional
arguments.  If you don't, someone reading your code might have to look up the
definition of a function with unfamiliar arguments to understand what you're
doing.

> ### Challenge
>
> Can you solve the following mathematical equations using R? 
>
> * 106 * 8
> * 1 + 2/3 * 4
> * (1+2)/3 * 4
> * (37+89)/12
> * $\sqrt{1654}$ 
> * 18\% of 97.65
> * log(100)
> * log(100, base = 10)
> * mean of: 34, 26, 65, 46, 27
>
> ```{r, answer=TRUE, results="markup", purl=FALSE}
>
>
> 106 * 8
> 1 + 2/3 * 4
> (1+2)/3 * 4
> (37+89)/12
> sqrt(1654)
> .18 * 97.65
## Notice the default is base e
> log(100) 
> log(100, base = 10)
> (34 + 26 + 65 + 46 + 27)/5
> ```

### Saving your code

Up to now, your code has been in the console. This is useful for quick queries
but not so helpful if you want to revisit your work for any reason.
A script can be opened by pressing <kbd>Ctrl</kbd> + <kbd>Shift</kbd> + 
<kbd>N</kbd>. 
It it wise to save your script file immediately. To do this press 
<kbd>Ctrl</kbd> + <kbd>S</kbd>. This will open a dialogue box where you 
can decide where to save your script file, and what to name it.
The `.R` file extension is added automatically and ensures your file
will open with RStudio.

Don't forget to save your work periodically by pressing <kbd>Ctrl</kbd> + 
<kbd>S</kbd>.


### Comments

The comment character in R is `#`, anything to the right of a `#` in a script
will be ignored by R. It is useful to leave notes and explanations in your
scripts.
RStudio makes it easy to comment or uncomment a paragraph: after selecting the
lines you  want to comment, press at the same time on your keyboard
<kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>C</kbd>. If you only want to comment
out one line, you can put the cursor at any location of that line (i.e. no need 
to select the whole line), then press <kbd>Ctrl</kbd> + <kbd>Shift</kbd> + 
<kbd>C</kbd>.

### Working with data 

Now that we have the basics of working with functions, let's work with some data. 

R can work with lots of different forms of data, like vectors, lists, and data.frames. We're going to start investigating the power of R using the data.frame. The curriculum we based this workshop on goes into more detail on these other data structures, and you can refer to it to learn more. 

```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup.R")
surveys <- read.csv("data_raw/portal_data_joined.csv")
suppressWarnings(surveys$date <- lubridate::ymd(paste(surveys$year,
                                                      surveys$month,
                                                      surveys$day,
                                                      sep = "-")))
```

We are going to use the R function `download.file()` to download the CSV file
that contains the survey data from Figshare, and we will use `read.csv()` to
load into memory the content of the CSV file as an object of class `data.frame`. 
Inside the download.file command, the first entry is a character string with the source URL 
("https://ndownloader.figshare.com/files/2292169"). This source URL downloads a CSV file from 
figshare. The text after the comma ("data_raw/portal_data_joined.csv") is the destination of the 
file on your local machine. You'll need to have a folder on your machine called "data_raw" where 
you'll download the file. So this command downloads a file from Figshare, names it 
"portal_data_joined.csv" and adds it to a preexisting folder named "data_raw".

```{r, eval=FALSE, purl=TRUE}
download.file(url = "https://ndownloader.figshare.com/files/2292169",
              destfile = "data_raw/portal_data_joined.csv")
```

You can navigate on your computer to this file and open it in Excel. Take a look at the data and familiarize yourself with it. If you are used to using Excel, think about how you would do things like filter out errors, count the number of records, fix missing data, compute means and medians, and make graphs. 

We'll read in our data using the `read.csv()` function.

First, let's find data_raw/portal_data_joined.csv on our computers and open it in Excel. Looks familiar, right?  Now, let's open it in R. 
```{r, results = 'hide', purl = FALSE}
surveys <- read.csv("data_raw/portal_data_joined.csv",
                    stringsAsFactors = FALSE)
```

Talk about stringsAsFactors as an optional argument that affects the functionality of the data. You'll usually want to use stringsAsFactors = F. Factors are one of the many data types that R can handle, but we're not going to discuss them in depth today. 


## Creating objects

Let's take a look at what we did there. 

`<-` is the assignment operator. It assigns values on the right to objects on
the left. We have created an object named "surveys". We can explore this object using different functions, or commands. 

```{r, purl = FALSE}
surveys          # prints the value 
```

For historical reasons, you can also use `=`
for assignments, but not in every context. Because of the
[slight](https://blog.revolutionanalytics.com/2008/12/use-equals-or-arrow-for-assignment.html)
[differences](https://r.789695.n4.nabble.com/Is-there-any-difference-between-and-tp878594p878598.html)
in syntax, it is good practice to always use `<-` for assignments.

In RStudio, typing <kbd>Alt</kbd> + <kbd>-</kbd> (push <kbd>Alt</kbd> at the
same time as the <kbd>-</kbd> key) will write ` <- ` in a single keystroke in a PC, while typing <kbd>Option</kbd> + <kbd>-</kbd> (push <kbd>Option</kbd> at the
same time as the <kbd>-</kbd> key) does the same in a Mac.

Objects can be given any name such as `x`, `current_temperature`, or
`subject_id`. You want your object names to be explicit and not too long. They
cannot start with a number (`2x` is not valid, but `x2` is). R is case sensitive
(e.g., `weight_kg` is different from `Weight_kg`). There are some names that
cannot be used because they are the names of fundamental functions in R (e.g.,
`if`, `else`, `for`, see
[here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Reserved.html)
for a complete list). In general, even if it's allowed, it's best to not use
other function names (e.g., `c`, `T`, `mean`, `data`, `df`, `weights`). If in
doubt, check the help to see if the name is already in use. It's also best to
avoid dots (`.`) within names. Many function names in R itself have them and
dots also have a special meaning (methods) in R and other programming languages.
To avoid confusion, don't include dots in names. It is also recommended to use nouns for object names, and
verbs for function names. It's important to be consistent in the styling of your
code (where you put spaces, how you name objects, etc.). Using a consistent
coding style makes your code clearer to read for your future self and your
collaborators. In R, three popular style guides are
[Google's](https://google.github.io/styleguide/Rguide.xml), [Jean
Fan's](https://jef.works/R-style-guide/) and the
[tidyverse's](https://style.tidyverse.org/). The tidyverse's is very
comprehensive and may seem overwhelming at first. You can install the
[**`lintr`**](https://github.com/jimhester/lintr) package to automatically check
for issues in the styling of your code.

We can also view the object in a prettier way, using commands

```{r, eval=FALSE, purl=FALSE}
## preview the data
View(surveys)
head(surveys)
tail(surveys)
dim(surveys)
```

The function `str()` provides an overview of the structure of an object and its
elements. It is a useful function when working with large and complex
objects:

```{r, results = 'hide', purl = FALSE}
## inspect the data
str(surveys)
```


## Inspecting `data.frame` Objects

We already saw how the functions `head()` and `str()` can be useful to check the
content and the structure of a data frame. Here is a non-exhaustive list of
functions to get a sense of the content/structure of the data. Let's try them out!

* Size:
    * `dim(surveys)` - returns a vector with the number of rows in the first element,
          and the number of columns as the second element (the **dim**ensions of
          the object)
    * `nrow(surveys)` - returns the number of rows
    * `ncol(surveys)` - returns the number of columns

* Content:
    * `head(surveys)` - shows the first 6 rows
    * `tail(surveys)` - shows the last 6 rows

* Names:
    * `names(surveys)` - returns the column names (synonym of `colnames()` for `data.frame`
	   objects)
    * `rownames(surveys)` - returns the row names

* Summary:
    * `str(surveys)` - structure of the object and information about the class, length and
	   content of  each column
    * `summary(surveys)` - summary statistics for each column

Note: most of these functions are "generic", they can be used on other types of
objects besides `data.frame`.


> ### Challenge
>
> Based on the output of `str(surveys)`, can you answer the following questions?
>
> * How many rows and how many columns are in this object?
> * What is `$` used to indicate? 
> * What is the difference in how R stores integer columns (int) versus character columns (chr)?
> * How does R store missing data? 
> * What is the class of the object `surveys`?
>
> ```{r, answer=TRUE, results="markup", purl=FALSE}
>
> str(surveys)
>
> ## * how many rows: 34786,  how many columns: 13
> ## * $ indicates column names
> ## * Integers do not have "" but characters do
> ## * Missing data are stored as NA
> ## * class: data.frame
>
> ```

> ### Further data exploration challenge
> 
> Using the functions listed for investigating a data.frame, can you answer the following questions?
> 
> * What is the last value in the column "genus"? 
> * What is the median weight? (hint: use `summary()`)
> * Is the median equal to the median in hindfoot length? 
> * Which column has the most missing data? 


